A primer to numerical simulations: The perihelion motion of Mercury

### 2018-04-17 tex/MercuryNumerical.tex
Added CHs reply and modifications to the PED ref

Numerical simulations play a key role in modern physics because they allow one to tackle some theoretical problems not accessible otherwise. This might be the case because there are too many particles participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
This paper introduces a project that allows one to demonstrate the power of numerical simulations.
On the example of the perihelion motion of the planet Mercury the students are supposed to learn about
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \vpython{}.
\end{itemize}
We are convinced that in order to excite students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
Their detailed study needs a deep understanding of Differential Geometry. For the course presented here, however, very little math is necessary, such that high school students vaguely familiar to vector calculus and derivatives will benefit from it.
This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017.

@@

A physical system is said to be understood, if the assumed forces acting on it lead to the observed trajectories.
In other words, we need to show that we can calculate the location in space of the object of interest at any future point in time, once the initial conditions are fixed properly.
If we can neglect the finite size of this object (and in particular its orientation in space), the location is parametrized by a single three vector $\vec r(t)$.
Additionally, we need to be able to calculate the objects velocity $\vec v(t)$ and acceleration $\vec a(t)$ in order to describe and control its dynamics; this will become clear below. The velocity describes a change of location over time which, for some infinitesimally small $\Delta t$, can be written as
\begin{equation}\label{eq:x_update}
\vec r(t+\Delta t) = \vec r(t) + \vec v(t) \Delta t + \ldots \ . \label{eq:vdef}
\end{equation}
Similarly, the acceleration describes a change in velocity, i.e.
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + \ldots\ . \label{eq:adef}
\end{eqnarray}
The dots in these expressions indicate additional terms that may be expressed in terms of higher powers in $\Delta t$. While they are needed in general, for sufficiently small $\Delta t$ they can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec r(t)}{\Delta t} =: \frac{d\vec r(t)}{d t}  = \dot{\vec  r}(t) \ ,
\end{equation}
where $\Delta \vec r(t)=\vec r(t+\Delta t)-\vec r(t)$ and we introduced a common short hand notation for time derivatives in the last expression.
Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec r(t)}{dt^2}  = \ddot{\vec r}(t) \ ,
\end{eqnarray}
where we introduced the second derivative in the second line.

It was Newton who observed that, if a body is at rest, it will remain at rest, and if it is in motion it will remain in motion at a constant velocity in a straight line, 
unless it is acted upon by some force --- this is known as Newton's first law.
Formulated differently: a force $\vec F$ expresses itself by changing the motion of some object.
This is quantified in Newton's second law
\begin{equation}
\vec F(\vec r, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation}
If the mass does not change~\footnote{%
	A well known example where $m$ does change with time is a rocket, whose mass decreases as fuel is burned.
} with time, this reduces to the well known
\begin{equation}
\vec F(\vec r) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{r}}(t) \ . \label{eq:newton2}
\end{equation}
Note that in general the force could depend also on the time or the velocity.
Here, we restrict ourselves to the case relevant for our example, where the force depends on the location only.
Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory by solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
Sometimes this requires some advanced knowledge of mathematics and sometimes no closed form solution exists.
However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
\begin{enumerate}
\item For a given time $t$, where $\vec r(t)$ and $\vec v(t)$ are known, use the force to calculate $\vec a(t) = \vec F(\vec r(t)) / m$, see Eq.~(\ref{eq:newton2}).
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec r(t+\Delta t)$.
\item Use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i) with $t\to t+\Delta t$.
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be known --- the  trajectories depend on these initial conditions~\footnote{%
	In general, a differential equation of $n^{\rm th}$ degree (where the highest derivative is order $n$) needs $n$ initial conditions specified.
	For $n=2$ those are often chosen as location and velocity at some starting time, but one may as well pick two locations at different times.%
}.

This procedure can only work if $\Delta t$ is sufficiently small.
One way to estimate whether $\Delta t$ is small enough is to verify whether the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec r(t)) |\Delta t\
\label{eq:check}
\end{equation}
holds. This follows from Eq.~(\ref{eq:vdef}) where the first term that we neglected reads $(1/2)a(t){(\Delta t)}^2$.
Eq.~({\ref{eq:check}}) also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since the time steps need to be small enough that all changes induced by the force get resolved.
Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive to the concrete value chosen for $\Delta t$ --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly.
This issue will be discussed in more detail below.

@@

If the starting values are chosen as advised in the previous section, the students should end up with a trajectory as depicted in Figure~\ref{fig:MercuryOrbit-a0-small}.
At this point one might ask students to vary $\Delta t$ and observe the effect of this on the orbit [see also Figure~\ref{fig:MercuryOrbit-a0-small-dt-large}].
To get the perihelion motion, the additional force term described in Eq.~(\ref{eq:newton_art}) has to be "turned on", e.g., by choosing either $\alpha\neq0$ or $\beta\neq0$
[Figure~\ref{fig:MercuryOrbit-a6-small}].
This is a good opportunity to let the students play with the size of $\alpha$ or $\beta$ and get a feeling for its impact on the trajectories. In particular
they should discover that values of $\alpha$ or $\beta$ larger than $10^5$ are necessary to get a visible effect.
For an even more enhanced perihelion motion it is advisable to not use the correct Mercury values but a more excentric trajectory by choosing, e.g., $r_{MS}(0)=6R_0$ [Figure~\ref{fig:MercuryOrbit-a6-big}].
\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a0-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/num-err.png}
		\caption{\label{fig:MercuryOrbit-a0-small-dt-large}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-big}}
	\end{subfigure}
	\captionsetup{singlelinecheck=off}
	\caption[]{\label{fig:MercuryOrbit}
		Different Mercury orbits for $\beta=0$ and
		\begin{itemize}
		\item[(a)] $\alpha=0$ and $\Delta t = \Delta t_0  / 20$;
		\item[(b)] $\alpha=0$ and $\Delta t = \Delta t_0  \times 2$, where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps;
		\item[(c)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
		\item[(d)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
		\end{itemize}
		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) /a_M(0)$ and the images are screenshots of the simulation (with modified colors).
		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "Mercury year" changes.
	}
\end{figure}

@@

>>>>

Numerical simulations play a key role in modern physics because they allow one to tackle some theoretical problems not accessible otherwise. This might be the case because there are too many particles participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
This paper introduces a project that allows one to demonstrate the power of numerical simulations.
On the example of the perihelion motion of the planet Mercury the students are supposed to learn about
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \vpython{}.
\end{itemize}
We are convinced that in order to excite students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
Their detailed study needs a deep understanding of Differential Geometry. For the course presented here, however, very little math is necessary, such that high school students vaguely familiar to vector calculus and derivatives will benefit from it.
This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017. As an example on how the course can be implemented in practice we describe our own
experiences with it in the second to last section.

@@

A  classical physics system is said to be understood, if the assumed forces acting on it lead to the observed trajectories.
In other words, we need to show that we can calculate the location in space of the object of interest at any future point in time, once the initial conditions are fixed properly.
If we can neglect the finite size of this object (and in particular its orientation in space), the location is parametrized by a single three vector $\vec r(t)$.
Additionally, we need to be able to calculate the objects velocity $\vec v(t)$ and acceleration $\vec a(t)$ in order to describe and control its dynamics; this will become clear below. The velocity describes a change of location over time which, for some infinitesimally small $\Delta t$, can be written as
\begin{equation}\label{eq:x_update}
\vec r(t+\Delta t) = \vec r(t) + \vec v(t) \Delta t + \ldots \ . \label{eq:vdef}
\end{equation}
Similarly, the acceleration describes a change in velocity, i.e.
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + \ldots\ . \label{eq:adef}
\end{eqnarray}
The dots in these expressions indicate additional terms that may be expressed in terms of higher powers in $\Delta t$. While they are needed in general, for sufficiently small $\Delta t$ they can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec r(t)}{\Delta t} =: \frac{d\vec r(t)}{d t}  = \dot{\vec  r}(t) \ ,
\end{equation}
where $\Delta \vec r(t)=\vec r(t+\Delta t)-\vec r(t)$ and we introduced a common short hand notation for time derivatives in the last expression.
Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t)  = \ddot{\vec r}(t) \ ,
\end{eqnarray}
where we introduced the second derivative in the last step.

The dynamics is controlled by Newton's second law
\begin{equation}
\vec F(\vec r, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation}
If the mass does not change~\footnote{%
	A well known example where $m$ does change with time is a rocket, whose mass decreases as fuel is burned.
} with time, this reduces to the well known
\begin{equation}
\vec F(\vec r) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{r}}(t) \ . \label{eq:newton2}
\end{equation}
Note that in general the force could depend also on the time or the velocity.
Here, we restrict ourselves to the case relevant for our example, where the force depends on the location only.
Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory by solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
Sometimes this requires some advanced knowledge of mathematics and sometimes no closed form solution exists.
However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
\begin{enumerate}
\item For a given time $t$, where $\vec r(t)$ and $\vec v(t)$ are known, use the force to calculate $\vec a(t) = \vec F(\vec r(t)) / m$, see Eq.~(\ref{eq:newton2}).
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec r(t+\Delta t)$.
\item Use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i) with $t\to t+\Delta t$.
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be fixed --- the  trajectories depend on these initial conditions~\footnote{%
	In general, a differential equation of $n^{\rm th}$ degree (where the highest derivative is order $n$) needs $n$ initial conditions specified.
	For $n=2$ those are often chosen as location and velocity at some starting time, but one may as well pick two locations at different times.%
}.
\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a0-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/num-err.png}
		\caption{\label{fig:MercuryOrbit-a0-small-dt-large}}
	\end{subfigure}
	\captionsetup{singlelinecheck=off}
	\caption[]{\label{fig:MercuryOrbit1}
		Different Mercury orbits for the purely Newtonian gravity and
		 $\Delta t = \Delta t_0  / 20$ for (a) and $\Delta t = \Delta t_0  \times 2$ for (b), 
		 where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps;
		Here, the reference time step is given by $\Delta t_0$ defined via Eq.~(\ref{eq:check}) as explained
		in the text.
		The images are screenshots of the simulation that is described below (with modified colors).
	}
\end{figure}

This procedure can only work if $\Delta t$ is sufficiently small.
One way to estimate whether $\Delta t$ is small enough is to verify whether the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec r(t)) |\Delta t\
\label{eq:check}
\end{equation}
holds. This follows from Eq.~(\ref{eq:vdef}) where the first term that was neglected reads $(1/2)a(t){(\Delta t)}^2$.
The impact of the size of $\Delta t$ on the simulation is illustrated in Fig.~\ref{fig:MercuryOrbit1} where
the force employed in Newtonian gravity as detailed in the next section. While for the
calculations for the left panel $\Delta t$
is chosen significantly below $\Delta t_0=2m |\vec v(0)| /   |\vec F(\vec r(0)) |$ ($cf.$ Eq.~(\ref{eq:check})), for the
right panel it was chosen twice as large. Accordingly only the trajectory in the left channel shows the characteristic
feature of a $1/r^2$--force of an ellipse fixed in space --- the trajectory in the right channel does not reproduce itself in successive revolutions.

Eq.~({\ref{eq:check}}) also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since the time steps need to be small enough that all changes induced by the force get resolved.
Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive to the concrete value chosen for $\Delta t$ --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly.
This issue will be discussed in more detail below.

@@

If the starting values are chosen as advised in the previous section, the students should end up with a trajectory as depicted in Figure~\ref{fig:MercuryOrbit-a0-small}.
At this point one might ask students to vary $\Delta t$ and observe the effect of this on the orbit [see also Figure~\ref{fig:MercuryOrbit-a0-small-dt-large}].
To get the perihelion motion, the additional force term described in Eq.~(\ref{eq:newton_art}) has to be "turned on", e.g., by choosing either $\alpha\neq0$ or $\beta\neq0$
[Figure~\ref{fig:MercuryOrbit-a6-small}].
This is a good opportunity to let the students play with the size of $\alpha$ or $\beta$ and get a feeling for its impact on the trajectories. In particular
they should discover that values of $\alpha$ or $\beta$ larger than $10^5$ are necessary to get a visible effect.
For an even more enhanced perihelion motion it is advisable to not use the correct Mercury values but a more excentric trajectory by choosing, e.g., $r_{MS}(0)=6R_0$ [Figure~\ref{fig:MercuryOrbit-a6-big}].
\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-big}}
	\end{subfigure}
	\captionsetup{singlelinecheck=off}
	\caption[]{\label{fig:MercuryOrbit}
		Different Mercury orbits for $\beta=0$ and
		\begin{itemize}
		\item[(a)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
		\item[(b)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
		\end{itemize}
		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) /a_M(0)$ and the images are screenshots of the simulation (with modified colors).
		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "Mercury year" changes.
	}
\end{figure}

@@

\section{Own experiences with the implementation of the course}

The idea that a numerical study of the perihelion motion of Mercury is an interesting subject for high
school students was taken from a project work of one of the authors (JH) presented at the 
M\"adchen-Gymnasium J\"ulich (a German high school) in 11$^{\rm th}$ grade in 2014. Based on this study
the course was developed as outlined in Sec.~\ref{sec:Numerical Implementation} to be used in 
the Sch\"ulerakademie Teilchenphysik" in 2015, which aims at  high school students from 10$^{\rm th}$ to 
13$^{\rm th}$ grade. This academy is financed by the German Research Society (DFG) as the outreach
branch
of a large research grant focussing on basic research in particle and nuclear physics.
One of the authors (CH) is acting as a Principal Investigator for the outreach
activities within this scheme. It runs biannually
at the Science Center Overbach in J\"ulich
for four days and has about 25 participants coming from various parts of Germany. 
The academy comprises lectures, a tour to the particle accelerator
COSY as well as the high performance computing facility at the Research Center J\"ulich as well
as one full day of practical work. The course presented here
 is offered as one option out of three on this day and was prepared
by a lecture to the whole group explaining the basics of numerical simulations. 
Given the setting the participating students
are highly motivated and eager to understand. Typically about one half of the participants chose to
work on the numerical project --- out of those about half had programmed a computer before while
the other half did not yet have any programming experience. Most of the participants had already
learned about derivatives in school. Given the very positive to enthusiastic 
feed backs we received from the participants we decided to make the course in the form presented
to a permanent part of the academy. It was used in the 2017 with the same success.

The students worked in pairs. Each group had an own laptop provided by us, where the necessary
software was preinstalled. Four of the authors (IH, CK, CM, JLW) acted as supervisors for
the students. 
The material covered in this one day is in most cases what is presented in Sec.~\ref{sec:Numerical Implementation}.
Only very few participants managed to start thinking about how to actually extract quantitatively the
perihelion motion --- not to speak of uncertainty estimates.
Once the basic problem was tackled some students deviated from the suggested path and for
instance studied the problem with one additional planet which was fine for us as well. This is
a very interesting problem by itself for there are parameter ranges where the classical three-body
system develops chaotic features.
 The reason why we included the
additional material in the paper is that we wanted to provide teachers with sufficient material to
probably even fill a working group outside the general curriculum running for a longer period.
